Estimating the Covariance Matrix

### 2024-02-05 src/explore/20240202-estcovmat.qmd
Finished initial simulation for estimating the covariance matrix.

## Proposals

### Proposal 1: REML

One potential solution would be to directly estimate $V$ and use some sort of
$\hat V$. We can do this with a REML estimate.

### Proposal 2: Orthogonal g-functions

@@

However, the parameters for $\tilde g_k$ and $k > 1$ are all regression
coefficients and the multiplier for $g_k$ is simply computing a variance.
Then to
achieve efficiency, we propose estimating the regression coefficients by running
the coresponding regression on all of the data points in $A_k$ as well as
computing the variance of $\tilde g_k$ within $A_k$.

@@

# Run from src/
source("R/data_generation.R")

@@

ortho_gls <- function(df, class = "basic") {

  # For class == 'basic', we consider the functions f_k to be originally,
  # f1 = X1, f2 = X2, f3 = Y.

@@

    # Get functions g_k
    g1 <- df$X1 / var(df$X1)
    df <- mutate(df, g1 = g1)
    df_1 <- filter(df, delta_11 == 1 | delta_10 == 1)
    t2 <- lm(X2 ~ g1, data = df_1)
    g2 <- df_1$X2 - t2$coefficients[2] * df_1$g1
    g2 <- g2 / var(g2)
    df_1 <- mutate(df_1, g2 = g2)
    df_11 <- filter(df_1, delta_11 == 1)
    t3 <- lm(Y ~ X1 + X2, data = df_11)
    g3 <- 
      df_11$Y - t3$coefficients[2] * df_11$g1 - t3$coefficients[3] * df_11$g2
    g3 <- g3 / var(g3)

    g_hat <- 
      c(mutate(df, tmp = delta_11 / pi_11 * g1) |> pull(tmp) |> sum(),
        mutate(df_1, tmp = delta_11 / pi_11 * g2) |> pull(tmp) |> sum(),
        mutate(df_11, tmp = delta_11 / pi_11 * g3) |> pull(tmp) |> sum(),
        mutate(df, tmp = delta_10 / pi_10 * g1) |> pull(tmp) |> sum(),
        mutate(df_1, tmp = delta_10 / pi_10 * g2) |> pull(tmp) |> sum(),
        mutate(df, tmp = delta_01 / pi_01 * g1) |> pull(tmp) |> sum(),
        # ERROR: This does not work!
        mutate(df_11, tmp = delta_01 / pi_01 * g3) |> pull(tmp) |> sum(),
        mutate(df, tmp = delta_00 / pi_00 * g1) |> pull(tmp) |> sum())

    g_hat <- g_hat / nrow(df)

  } else {
    stop("Only class == 'basic' has been implemented.")
  }
}

@@

<!---# Compare with: Best case in monotone data --->
## Monotone Case

# Variance Estimation

@@

Did it work?

>>>>

## Proposal

Instead of using the known covariance matrix $V$, we estimate it instead.
Suppose that we observe variables $Z = (X_{m_1}, Y_{m_2})$ with an objective of
estimating $\theta = E[g(Y_{m_2})$ for some known function $g$ such that the
variables $Z$ are subject to missingness. We assume that there are $R$ unique
combinations of observed variables including a fully observed case. We can index
the combinations of observed variables by $r$ and assume that the fully observed
case occurs at $r = 1$. Let $G_r(Z)$ be the variables that are observed at a
particular value of $r$. We can choose a sequence of functions $f_1, \dots, f_K$
that we want to estimate. Each $f_k$ is assumed to be a function of a subset of
$Z$ and it makes sense to assume (since these are chosen by the analyst) that
there is at least one $f_k$ for each combination of observed variables $G_r(Z)$.
Let $A_k$ be the sets of observed variable combinations that can be evaluate by
the function $f_k$. If $f_k$ can be evaluated by the observed variables
$G_r(Z)$, this will consist of all of the combinations of variables
$r'$ such that $G_r(Z) \subseteq G_r'(Z)$. Because we assume that $G_1(Z) = Z$,
$A_k$ is always non-empty.

To estimate the covariance matrix $V$, we estimate the covariance between 
$f_{k_1}$ and $f_{k_2}$ directly by computing the estimated covariance on 
$A_{k_1} \cap A_{k_2}$.

<!---# This did not work because the orthogonalization created variables that we
needed to impute
### Proposal 2: Orthogonal g-functions

@@

However, the parameters for $\tilde g_k$ and $k > 1$ are all regression
coefficients and the multiplier for $g_k$ is simply computing a variance.
Then to
achieve efficiency, we propose estimating the regression coefficients by running
the coresponding regression on all of the data points in $A_k$ as well as
computing the variance of $\tilde g_k$ within $A_k$.
--->

@@

# Run from src/ or src/explore/
source("R/data_generation.R")

@@

est_var_gls <- function(df, class = "basic") {

@@

    z_mat <- matrix(c(1, 0, 0,
                      0, 1, 0,
                      0, 0, 1,
                      1, 0, 0,
                      0, 1, 0,
                      1, 0, 0,
                      0, 0, 1,
                      1, 0, 0), ncol = 3, byrow = TRUE)

    # Estimate Covariance Matrix
    df_1p <- filter(df, delta_11 == 1 | delta_10 == 1)
    df_p1 <- filter(df, delta_11 == 1 | delta_01 == 1)
    df_11 <- filter(df, delta_11 == 1)
    v_x1 <- var(df$X1)
    v_x2 <- var(df_1p$X2)
    v_y <- var(df_p1$Y)
    cov_x1x2 <- cov(df_1p$X1, df_1p$X2)
    cov_x1y <- cov(df_p1$X1, df_p1$Y)
    cov_x2y <- cov(df_11$X2, df_11$Y)

    v_11 <- 1 / df$pi_11[1] *
      matrix(c(v_x1, cov_x1x2, cov_x1y,
               cov_x1x2, v_x2, cov_x2y,
               cov_x1y, cov_x2y, v_y), nrow = 3)

    v_10 <- 1 / df$pi_10[1] * matrix(c(v_x1, cov_x1x2, cov_x1x2, v_x2), nrow = 2)
    v_01 <- 1 / df$pi_01[1] * matrix(c(v_x1, cov_x1y, cov_x1y, v_y), nrow = 2)
    v_00 <- 1 / df$pi_00[1] * v_x1

    return(as.numeric(t(c(0, 0, 1) %*% ghat)))
  } else {
    stop("Only class == 'basic' has been implemented.")
  }
}

@@

```{r}
#| label: MCMC
#| echo: false
#| tbl-cap: |
#|   Results from simulations study with independent equally sized segments
#|   $A_{11}$, $A_{10}$, $A_{01}$, and $A_{00}$ all of size $n = 250$. In this
#|   simulation we have the true mean of $Y_2$ equal to $\mu = 5$ and the covariance
#|   between $e_1$ and $e_2$ is $\rho = 0.5$. The goal is to estimate 
#|   $E[Y_2] = \mu$. For the GLS estimation, we use the
#|   estimated covariance matrix $\hat V$ with 
#|   f-functions $f_1 = X_1$, $f_2 = X_2$ and $f_3 = Y$.

clust <- parallel::makeCluster(min(parallelly::availableCores() - 2, 100))
registerDoParallel(clust)

mc_theta <-
  foreach(iter = 1:B,
          .options.RNG = 1,
          .packages = c("dplyr", "stringr")) %dorng% {

    # Generate Data
    df <- gen_simple_data(obs_seg, mu = true_mu, rho = cov_e1e2)

    # Estimators
    oracle <- mean(df$Y)
    cc <- df |>
      filter(delta_11 == 1) |>
      pull(Y) |>
      mean()
    ipw <- df |>
      mutate(tmp = delta_11 / pi_11 * Y) |>
      pull(tmp) |>
      mean()

    gls <- gls_est(df, rho = cov_e1e2)
    gls_ev <- est_var_gls(df)

    return(tibble(Oracle = oracle,
                  CC = cc,
                  IPW = ipw,
                  GLS = gls,
                  GLSEstVar = gls_ev
    ))

  } |>
  bind_rows()

stopCluster(clust)

# Analyze Results
mc_theta |>
  summarize(
    across(everything(),
    list(Bias = ~mean(.) - true_mu,
         SD = ~sd(.),
         Tstat = ~(mean(.) - true_mu) / sqrt(var(.) / B)),
    .names = "{.fn}_{.col}")) |> 
  tidyr::pivot_longer(cols = everything(),
                      names_to = c(".value", "Algorithm"),
                      names_pattern = "(.*)_(.*)") |>
  mutate(Pval = pt(-abs(Tstat), df = B)) |>
  knitr::kable(digits = 3)

<!---# Compare with: Best case in monotone data 
## Monotone Case--->

# Optimal choice of g-functions

TODO for the next set of simulations.

@@

Did it work? Yes, estimating the covariance matrix did not produce a loss in
efficiency.
